The Impact of Disk Winds on the Optical Spectra of Cataclysmic Variables

### 2014-10-13 draft.tex
updated to ksl version

\begin{abstract}
High-state non-magnetic cataclysmic variables (CVs) 
exhibit strong hydrogen and helium recombination lines in their optical 
spectra. These lines can be single peaked, and may be formed 
in the same bipolar outflow that is responsible for the
blue-shifted absorption lines seen in the ultraviolet (UV).
Here, we present results obtained by incorporating a `macro-atom' treatment into
our Monte Carlo radiative transfer code, which was
originally used to synthesize the UV spectra of disk-dominated, high-state CVs 
with a simple biconical wind. 
Using an identical model, we show that the same wind leaves a noticeable imprint 
on the optical spectrum, producing \ha\ and \heiiopt\ as well as
enhanced emission in the Ly-$\alpha$ and \heiiuv\ UV lines. 
% In addition, our improved treatment of recombination means that the bound-free continuum
% emission is significant, but does not yet overcome the intrinsic absorption edge
% present in the stellar atmosphere models used to treat the accretion disk.
We find that a boundary layer is not essential for the production
of He II recombination lines.
%  but also explore the effect of boundary layers
% with temperatures of $80-200$kK on the ionization state of the wind
% and resulting spectrum.
A short exploration of a 3-variable kinematic parameter space leads to models with higher densities 
that are able to produce single-peaked line emission, and also generate sufficient recombination emission
to fill in the Balmer jump photoabsorption edge intrinsic to our disk atmosphere spectra.
Finally, we present a synthetic spectrum of an improved optical model in 
and out of eclipse, which shows a striking similarity with 
the optical spectrum of the high inclination
Nova-like variable RW Tri. The densities near the disk plane in this model
are such that \civfull\ has a large collisionally excited component, in 
contrast to observations.
% We find that 
% a boundary layer is not essential to produce significant recombination to 
% He II, and if one is present we favour temperatures of $<10^5$K to avoid
% overionizing the UV resonance line species such as CIV. We also conduct Zanstra method
% calculations and estimate that one would infer a boundary layer temperature of around 80,000K
% from our benchmark calculation, despite it not being present in the model itself. This suggests that previous calculations
%done using this method are strongly dependent on both the assumed geometry and disk
%spectrum.
%We also note that
%the presence of a boundary layer becomes more noticeable in lower accretion rate systems.
\end{abstract}

@@

Cataclysmic variables (CVs) are systems in which a white dwarf accretes matter from a donor
star via Roche-lobe overflow. In non-magnetic, disk-dominated systems this accretion
is mediated by an accretion disk which forms around the white dwarf. 
Nova-like variables (NLs) are a particular subclass of objects in which the accretion disk
is in a permanent state of relatively high accretion rate 
($\dot{M} \sim 10^{-8}$~M$_{\odot}$~yr$^{-1}$),
making them the perfect laboratories to test the widely-used 
`$\alpha$-disk' model proposed by 
\cite{shakurasunyaev1973}.

For over three decades, it has been known that winds emanating from the accretion disk
are important in shaping the ultraviolet (UV) spectra of high-state CVs (Heap 1978), 
the most spectacular evidence being the P-Cygni like profiles of resonance lines such as 
\civfull\ (see e.g. Cordova \& Mason 1982\nocite{cordova1982}).
Despite this, in many types of systems the impact of winds on observational
features is essentially unknown, and many theoretical questions relating to their origin 
and driving mechanism remain unanswered.

While the effect of the wind on the UV is at least partially clear, it may also have more subtle 
effects across the spectral range. It has been proposed that an accretion disk wind
can provide a mechanism for single-peaked emission lines (Murray \& Chiang 1996), 
such as those seen in Nova-like variables and, especially, the SW Sex stars \citep{HSK86, DR95}.
It is particularly telling that single-peaked 
lines in high-inclination systems can be seen even in eclipse, suggesting 
that a spatially extensive wind may be the source of emission (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence of the influence of winds in the optical can occasionally 
be occasionally seen in P-Cygni like profiles observed in lines such as \ha\ 
and He \textsc{i} $\lambda5876$ \citep{RN98}.

@@

It has also been proposed
that recombination continuum emission from a disk wind or `transition region'
could be responsible for filling in the Balmer absorption edge  \citep{KLWB98}. 
This provides a potential explanation for
the absence of a pronounced `Balmer jump' in optical spectra of CVs, 
which has never been adequately produced by using stellar atmospheres 
to model the accretion disk (see e.g. Haug 1987)\nocite{haug1987}. 
However, it is also possible that accretion disks do not possess
and intrinsic deep Balmer absorption edges.
This depends to the optical depth at the edge
and can change due to decrease due to less temperature variation with height
near the disk photosphere (see e.g. Shaviv et al. 1991; 
but also more recent work by Idan et al. 2010).
\nocite{idanshaviv2010} \nocite{shaviv1991}

@@

Here, we present work using the same Monte Carlo radiative transfer code as LK02, 
but incorporating additional `macro-atom' techniques suggested by Lucy (2002, 2003). 
Our aim is to model high-state CVs over a wide spectral range.
In particular, we hope to assess the potential contribution
to the optical spectrum from a disk wind model which preserves the UV features.
It will then be possible to see how closely
our models resemble the spectra in figure 1 as we change wind parameters.
We will investigate whether a wind can produce significant
recombination emission lines and whether the recombination
continuum emission is sufficient to mask the Balmer jump from a disk atmosphere.
We will also learn if the mechanism for single-peaked line emission
proposed by \cite{MC96} works with plausible model parameters.
In section 2 we present the code and provide a brief description of macro-atoms 
and the radiative transfer methods we employ. 
We describe the kinematics and geometry of our biconical wind model in section 3.
In section 4, we present a benchmark model, which is explored on in section 5 
with a small-scale parameter search. In section 6, we present a
model which better reproduces the optical spectrum of a NL, and, 
finally, in section 7, we summarise our findings.

@@

\py is a Monte Carlo (MC) radiative transfer code which uses
the Sobolev approximation to treat line transfer. It operates 
in two distinct cycles. First, the ionization state, level populations
and temperature structure are calculated. This is done by
flying MC energy quanta (`photons') through a pre-defined wind geometry and recording
the relevant estimators, before choosing  a new electron temperature 
in an attempt to balance heating and cooling in the plasma.
This process is repeated until the code converges on a 
solution. Second, the spectrum is synthesized by tracking
the energy packets through the wind and computing the 
emergent spectra at a number of user-specified viewing angles.

The code is capable of operating in a number of different
regimes, both in terms of the scale of the system, and the 
shape and characteristic energies of the illuminating spectrum. 
It was originally developed by LK02 in order to model the UV spectra
of CVs with a simple biconical disk wind model.. 
Higginbottom et al.\nocite{higginbottom2013} (2013; hereafter H13) used the same code to
present a benchmark model for broad absorption line (BAL) QSOs.
As part of this effort, the ionization scheme was improved so that 
the code is now capable of treating arbritrary SEDs.
%the power-law X-ray spectrum exhibited by AGN.
% This involved using a spectral model the mean intensity in a cell
% computed from MC estimators for, instead of using a dilute blackbody
% approximation. Compton processes
% were also incorporated into the heating and cooling balance.
In addition, Sim et al. (2005; hereafter SDL05)\nocite{simmacro2005} modeled
Brackett and Pfund line profiles in young-stellar objects (YSOs) using \textsc{Python,} which included
implementing a `macro-atom' mode in order to correctly treat recombination lines.

@@

Previously to SDL05, \py used a two-level atom approximation \cite[see e.g.][]{mihalas}. This approximation is 
fast, and works well as a first approximation when treating so-called `resonance lines' (such as \civfull) 
in which the excited electron is strongly coupled to the ground state.
However, in a number of situations, for example, when a level is primarily populated from above, then
this becomes a poor representation. 
To reproduce recombination lines and lines in which the lower level is not the ground state, 
it is therefore important to use an improved line transfer technique and level population solver.

@@

Whereas SDL05 used \py itself to apply the macro-atom 
technique to a Hydrogen-only model, we
adopt a hybrid ionization and excitation scheme. 
Species can be treated as either full macro-atoms
or using the two-level atom approximation.
In the latter case, they still follow the indivisible packet constraint.
This allows one to preserve the fast treatment of resonance lines while providing 
an improved representation of processes with more complex
transition paths such as recombination. This is important when modeling features 
such as the Balmer lines and also recombination {\sl continuum} emission.

@@

\subsection{Atomic Data}
Here we use the same atomic data as LK02 and updated by H13.
We have obtained the additional level and line information required to treat He as a macro-atom 
from \textsc{Topbase} \citep{topbase2005}. We treat H as a 20-level atom as in SDL05, where 
each level is defined by the principal quantum number, $n$, which is appropriate due to
the degeneracy of each level in the H atom. He~\textsc{II} is treated in much the same way,
but with 10 levels. However, He~\textsc{I} has larger energy differentials between different l subshells
and triplet and singlet states. Thus, we still include levels up to $n=10$ but explicitly 
treat the $l$ and $s$ sub-orbitals as distinct levels instead of assuming they are perfectly `l-mixed',
This also allows us to model singlet and 
triplet lines, which are seen in optical CV spectra \citep[e.g.][]{dhillon1996}.

@@

\py has previously been tested against a number of radiative transfer and
photoionization codes. LK02 and H13 conducted comparisons of ionization balance 
with \cld \citep{cloudy2013}, demonstrating excellent agreement. 
We have also carried out comparisons
of ionization and spectral synthesis with the supernova code \textsc{Tardis.} \tar is described by
\cite{kerzendorfsim}, and the spectral comparisons can be found therein.
In addition to these code validation efforts, we have carried out tests of
the macro atom scheme in \py to verify that it does indeed solve for level populations correctly
and produces the expected level emissivities. Figure~\ref{tests} shows good agreement with 
l-mixed Balmer decrement calculations in the Case B limit by \cite{seaton1959}, and 
also compares He \textsc{i} level populations with \textsc{Tardis.}

@@

Here we describe some of the features of our biconical wind model.

@@

For our kinematic wind model we use 
the prescription outlined by Schlosman \& Vitello (1993; hereafter SV93). A schematic is shown
in figure~\ref{cartoon}. A smooth, biconical
disk wind rises from the accretion disk between radii $r_{min}$ and $r_{max}$. The covering fraction of the wind is 
also controlled by the opening angles of the wind, $\theta_{min}$ and $\theta_{max}$, and the launch angle of
the other streamlines is given by

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/acc_law.eps}
\caption{
The velocity law in our model for various values of
the acceleration exponent, $\alpha$.
}
\label{acc_law}
\end{figure}

@@

Photons originate from the accretion disk and a central WD in our model. It is also possible to include
a boundary layer with user-defined temperature and emitting area. Emission from the wind is dealt with
by reprocessing of the indivisible energy packets originally produced in the disk or central WD.

@@

\py has has some flexibility when treating the accretion disk as a source of photons. 
The disk is broken down into annuli 
such that each annulus contributes an equal amount to the bolometric luminosity. 
We adopt the temperature profile of a standard \cite{shakurasunyaev1973} $\alpha$-disk.
%shown in figure 1.
An annulus can then
be treated either as a blackbody with the corresponding effective temperature or as a stellar atmosphere model
with the appropriate surface gravity and effective temperature. 
Both of these methods have limitations. It is known
from studies of CV accretion disks that they exhibit absorption cores 
formed in an optically thick disk atmosphere.
This can impact the overall spectrum even if a wind is present 
\citep[see e.g.][]{dhillon1996}.
However, the stellar atmosphere models
have had difficulties reproducing the required spectral shape \citep{wade1988}, 
and there are inconsistencies when using a 
stellar atmosphere beneath a dense
wind region that itself goes some way to simulating the disk atmosphere. 
In addition, \cite{linnell2010} found that deviations from the predicted 
$T\propto r^{-3/4}$ temperature profile are required to fit the data
in certain cases.

We hope that future simulations, coupled with a comparison to data, will 
actually address some of these problems, particularly since
the filtering effect of the wind may cause noticeable effects
on the colour of the emergent spectrum \citep{hassall}. 
Here, we use a blackbody illuminating spectrum
for the ionization cycles (see figure~\ref{sed}) and to compute our MC estimators.
During the spectral synthesis stage of the simulation we use stellar atmosphere models.
This is partly because they provide the most pessimistic prediction in terms of whether recombination emission in the 
wind can fill in the Balmer jump, due to their intrinsic strong absorption beyond the Balmer edge.
For low temperatures, we use stellar atmospheres computed 
by Kurucz \citep{kurucz1991}, and for $T>50,000$~K we use models computed with 
\textsc{TLUSTY} \citep{tlusty}. The input spectra are created from these model
atmospheres using \textsc{Synspec}\footnote{http://nova.astro.umd.edu/Synspec43/synspec.html}.
%The temperature profile of our disk,
%with $\dot{M}_{acc} = 10^{-8}~M_{\odot}yr^{-1}$ can be seen in figure~\ref{t_profile}.
% more detail on why we use this hodge podge of atmosphere models?

\subsubsection{Boundary Layer}
The existence of a boundary layer in CVs arises naturally from the boundary condition
of the \cite{shakurasunyaev1973} model, since the material at the inner disk
edge must lose angular momentum in order to be deposited on
the surface of the compact object \cite[see e.g.][]{lyndenbell1974}.
Boundary layers are a source of
high energy photons which may be responsible for the photoionization of 
He~\textsc{ii}, so a disk wind can therefore be an important diagnostic
tool for inferring boundary layer luminosities and temperatures. 
The luminosity of the boundary layer, $L_{bl}$, is related to the accretion luminosity
by

@@

It is possible to include a boundary layer in our model which emits 
as a blackbody with a user-specified effective temperature and luminsoity.
Boundary layer photons are then produced around the spherical surface of the 
white dwarf. In our benchmark model, we follow LK02 in setting the boundary layer
luminosity to zero. We investigate the effect
of boundary layers with temperatures between $80{\rm kK}$ and $200{\rm kK}$ 
on the ionization state of the wind and resulting
spectrum in section 5.

@@

Our aim here is not
to provide a `best fit' model to match any particular dataset.
Rather, we take
a successful UV wind model as a starting point
and examine its effect 
when one utilises the better treatment of lines provided
by the macro-atom scheme.
The model parameters are shown in table 1. 
they are identical to LK02, and were chosen both to reflect typical
values for NLs and reproduce realistic UV line profiles.
% The specific model 
% parameters are summarised in table 1, and follow LK02. 
% They are motivated by a need to agree with typical observed values 
% for NLs and the values required to reproduce the UV line profile

@@

Figure~\ref{spec} shows optical synthetic spectra for the benchmark 
model for five different inclination angles ranging from $10^\circ$ to $80^\circ$ 
%5 different sightlines.  
The model produces some lines which are strong enough to overcome
the absorption cores lines from the disk. These lines are generally double peaked, 
as they are formed near the base of the wind where rotational velocity
can still dominate over the slowly increasing poloidal velocity. The line 
opacity is not yet high enough for the poloidal velocity gradient to 
cause single-peaked profiles. 
We also see a general trend from absorption to emission in spectral lines
with increasing inclination, as one would expect from our wind geometry.
This trend is observed in CVs and can be seen clearly in figure 1. 
% Note, however,
% that in our models the \heiiuv\ line does not exhibit this behaviour, which
% is somewhat surprising. {\bf NOTE: we need to understand this!}.

@@

In addition to examining synthetic spectra, we are also able to 
analyse the physical properties of the outflow.
Figure~\ref{wind} shows various physical properties
of the wind geometry. The ionization structure of the wind
is very similar to the original LK02 model, as one might expect.
He is mostly ionized throughout the wind except for a small region near
the disk which is shielded from the photons produced by the hot inner disk.
C\textsc{iv} is the dominant Carbon ion throughout the wind, allowing
for an absorbing column which covers a large range of velocities and is thus
sufficient to produce the required broad, deep and blue-shifted absorption lines
we expect to see in a large fraction of nova-likes.

@@

Clearly, this particular model cannot explain all
the features seen in the optical spectra of nova-like variables. However,
the wind has a significant effect on the observed spectral features. 
It cannot be ignored as a source of line and continuum emission and as a reprocessing `filter'
of the underlying disk spectral energy distribution (SED).

>>>>

\begin{abstract}
High-state non-magnetic cataclysmic variables (CVs) 
exhibit strong hydrogen and helium recombination lines in their optical 
spectra. These lines can be single peaked.  The Balmer jump observed in such systems
is less pronounced than one would expect if the continuum in these systems arises solely from disk emission
calculated by summing appropriately weighted stellar atmospheres.  Here we investigate the possibility that
both of these effects are due to emission from the bipolar outflows that are known to be 
  responsible for the
blue-shifted absorption lines seen in the ultraviolet (UV).  For this study, we used an improved version of
the Monte Carlo radiative transfer code that we used previously to show qualitatively that the UV spectra of disk-dominated,
high state CVs could be reproduced from a relatively simple parameterization of a biconical wind with a mass loss rate about one-tenth of the mass accretion rate.
%Here, we present results obtained by incorporating a `macro-atom' treatment into our Monte Carlo radiative transfer code, which was originally used to synthesize the UV spectra of disk-dominated, high-state CVs  with a simple biconical wind. 
We find that the same parameterized model leaves a noticeable imprint 
on the optical spectrum, producing \ha\ and \heiiopt\ as well as
enhanced emission in the Ly-$\alpha$ and \heiiuv\ UV lines. 
% In addition, our improved treatment of recombination means that the bound-free continuum
% emission is significant, but does not yet overcome the intrinsic absorption edge
% present in the stellar atmosphere models used to treat the accretion disk.
A boundary layer is not essential for the production
of He II recombination lines.  However, all of the optical lines produced from our original parameterization are double-peaked, and the effect on the Balmer line region is modest.
%  but also explore the effect of boundary layers
% with temperatures of $80-200$kK on the ionization state of the wind
% and resulting spectrum.
Based on  a short exploration of a 3-variable kinematic parameter space, we conclude that  models with higher densities are needed 
 to produce single-peaked line emission, and to generate sufficient recombination emission
to fill in the Balmer jump photoabsorption edge intrinsic to our disk atmosphere spectra.
The  synthetic spectra using the revised set of parameters  shows  striking similarity with 
the optical spectrum of the high inclination
Nova-like variable RW Tri, both in and out of eclipse. The densities near the disk plane in this model
are such that \civfull\ has a large collisionally excited component, in 
contrast to observations. {\bf ksl - I am not sure what the last sentence here means.  How do we know from observations that CIV is collisionally excited}.   Nevertheless, it seems 
quite possible that models with a different parameterization of the wind will ultimately be able to reproduce the spectra of high state, non-magnetic CVs from the UV through the visible.
% We find that 
% a boundary layer is not essential to produce significant recombination to 
% He II, and if one is present we favour temperatures of $<10^5$K to avoid
% overionizing the UV resonance line species such as CIV. We also conduct Zanstra method
% calculations and estimate that one would infer a boundary layer temperature of around 80,000K
% from our benchmark calculation, despite it not being present in the model itself. This suggests that previous calculations
%done using this method are strongly dependent on both the assumed geometry and disk
%spectrum.
%We also note that
%the presence of a boundary layer becomes more noticeable in lower accretion rate systems.
\end{abstract}

@@

Cataclysmic variables (CVs) are systems in which a white dwarf accretes matter from a donor
star via Roche-lobe overflow. In non-magnetic systems this accretion
is mediated by an accretion disk which forms around the white dwarf. 
Nova-like variables (NLs) are a particular subclass of objects in which the accretion disk
is in a permanent state of relatively high accretion rate 
($\dot{M} \sim 10^{-8}$~M$_{\odot}$~yr$^{-1}$).  This makes nova-like variables the obviousl
laboratory for studying the properties of steady-state accretion disks.
%making them the perfect laboratories to test the widely-used`$\alpha$-disk' model proposed by \cite{shakurasunyaev1973}.

For over three decades, it has been known that winds emanating from the accretion disk
are important in shaping the ultraviolet (UV) spectra of high-state CVs (Heap 1978), 
the most spectacular evidence being the P-Cygni like profiles of resonance lines such as 
\civfull\ (see e.g. Cordova \& Mason 1982\nocite{cordova1982}).  As discussed below, it is now possible to produce synthetic spectra 
of CVs with winds that are qualitatively very similar to those observed.  However, less is known about the 
the effects of the wind on the optical spectra of high state CVs, which are characterized by a relatively featureless 
continuum and  emission from the
Balmer lines of H, which in many cases are single peaked, especially, the SW Sex stars \citep{HSK86, DR95}, unlike the double-peaked lines one might expect from a disk.   Some work has been carried out.
% Despite this, in many types of systems the impact of winds on observational features is essentially unknown, and many theoretical questions relating to their origin  and driving mechanism remain unanswered.
Murray \& Chiang (1996) have argued that the single peaked emission lines arise due to radiative transfer effects in the wnd
% While the effect of the wind on the UV is at least partially clear, it may also have more subtle  effects across the spectral range. It has been proposed that an accretion disk wind can provide a mechanism for single-peaked emission lines (Murray \& Chiang 1996),  such as those seen in Nova-like variables and, especially, the SW Sex stars \citep{HSK86, DR95}.
It is particularly telling that single-peaked 
lines in high-inclination systems can be seen even in eclipse, suggesting 
that a spatially extensive wind may be the source of emission (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence of the influence of winds in the optical can occasionally 
be occasionally seen in P-Cygni like profiles observed in lines such as \ha\ 
and He \textsc{i} $\lambda5876$ \citep{RN98}.

@@

The continuum spectra of high disk-dominated CVs is generally assumed to be
dominated by emission from the diskand and is often modeled in terms of a set of stellar atmospheres
and/or disk models ({bf reference some of Linnells papers here}), using Hubeny's TLUSTY code {\cite{})
The optical continua observed from  high-state, disk-dominated CVs lack pronounced ``Balmer jumps'', 
spectra simulated from summed atmospheres.   One possible explanation for this problem is that disk atmospheres calculated
with TLUSTY simply do not represent the physics of disks properly ((see e.g. Shaviv et al. 1991; 
but also more recent work by Idan et al. 2010).
\nocite{idanshaviv2010} \nocite{shaviv1991}.   An altenative suggested by \cite{KLWB98} is that recombination continuum emission from a disk wind or `transition region'
fills  in the Balmer absorption edge.

%It has also been proposed that recombination continuum emission from a disk wind or `transition region'could be responsible for filling in the Balmer absorption edge  \citep{KLWB98}. This provides a potential explanation for the absence of a pronounced `Balmer jump' in optical spectra of CVs,  which has never been adequately produced by using stellar atmospheres to model the accretion disk (see e.g. Haug 1987)\nocite{haug1987}. However, it is also possible that accretion disks do not possessand intrinsic deep Balmer absorption edges. This depends to the optical depth at the edge and can change due to decrease due to less temperature variation with height  near the disk photosphere (see e.g. Shaviv et al. 1991;  but also more recent work by Idan et al. 2010).\nocite{idanshaviv2010} \nocite{shaviv1991}

@@

Here, we present work using the same Monte Carlo radiative transfer code as LK02, 
but incorporating additional `macro-atom' techniques suggested by Lucy (2002, 2003). 
Our aim is to model high-state CVs over a wide spectral range.
In particular, we hope to assess the potential contribution
to the optical spectrum from a disk wind model which preserves the UV features.
It will then be possible to see how closely
our models resemble the spectra in figure 1 as we change wind parameters.
We will investigate whether a wind can produce significant
recombination emission lines and whether the recombination
continuum emission is sufficient to mask the Balmer jump from a disk atmosphere.
We will also learn if the mechanism for single-peaked line emission
identified by \cite{MC96} operates  in  parameterized wind models capable of reproducing 
other spectral features of observed CVs.  The remainder of this report is organized as followd
In section 2 we present the code and provide a brief description of macro-atoms 
and the radiative transfer methods we employ.  In section 3, 
We describe the kinematics and geometry of our biconical wind model.
In section 4, we present spectra simulated from the benchmark model employed by LK02, and in section 5, we carry out 
aparameter search to determine how spectra change as a function of several of the parmeters in the model. In section 6, we present a
model which better reproduces the optical spectrum of a NL, and, 
finally, in section 7, we summarise our findings.

@@

\py is a Monte Carlo (MC) radiative transfer code which uses
the Sobolev approximation to treat line transfer. It operates 
in two distinct cycles. First, the ionization state, level populations
and temperature structure are calculated. This is done by
flying MC energy quanta (`photons') through a pre-defined wind geometry and recording
estimators that characterize the radiation field in each cell of the grid, before choosing  a new electron temperature 
that more closely balances heating and cooling in the plasma. Using the new electron temperature and the radiative estimators, we then
revise the ionization state of the wind.  
This process is repeated  a converged
solution is reached with heating and cooling balanced throughout the wind. Second, the spectrum over the desired spectral range is synthesized by tracking
the energy packets through the wind and computing the 
emergent spectra at a number of user-specified viewing angles.

\py is designed to operate in a number of different
regimes, both in terms of the scale of the system, and the 
shape and characteristic energies of the illuminating spectrum. 
It was originally developed by LK02 in order to model the UV spectra
of CVs with a simple biconical disk wind model.. 
Higginbottom et al.\nocite{higginbottom2013} (2013; hereafter H13) used the same code to
present a benchmark model for broad absorption line (BAL) QSOs.
As part of this effort, the ionization scheme was improved so that 
the code is now capable of treating arbritrary SEDs.
%the power-law X-ray spectrum exhibited by AGN.
% This involved using a spectral model the mean intensity in a cell
% computed from MC estimators for, instead of using a dilute blackbody
% approximation. Compton processes
% were also incorporated into the heating and cooling balance.
In addition, Sim et al. (2005; hereafter SDL05)\nocite{simmacro2005} used the code to model
Brackett and Pfund line profiles of H in young-stellar objects (YSOs) using \textsc{Python,} and as part of this effort 
implemented a `macro-atom' mode  in order to correctly treat recombination lines of H.

@@

Previously to SDL05, \py used a two-level atom approximation \cite[see e.g.][]{mihalas}. This approximation is 
fast, and works well as a first approximation when treating  `resonance lines' (such as \civfull) 
in which the excited electron is strongly coupled to the ground state.
However, in a number of situations, for example, when a level is primarily populated from above, then
this becomes a poor representation. 
To reproduce recombination lines and lines in which the lower level is not the ground state, 
it is therefore important to use an improved line transfer technique and level population solver.

@@

Whereas SDL05 used \py itself to apply the macro-atom 
technique to a Hydrogen-only model, we
adopt a hybrid ionization and excitation scheme to model the spectra of CVs. 
Atomic pecies can be treated as either full macro-atoms
or using the two-level atom approximation.  This is  important in our multi-dimensional code to balance the need to calculate certain transitions in detail and the desire to minimize the set of
computational resources required.  For CVs, this allows us 
%In the latter case, they still follow the indivisible packet constraint.This allows one 
to preserve the fast treatment of the UV resonance lines while providing 
an improved representation of processes with more complex
transition paths such as recombination, H and He. %This is important when modeling features  such as the Balmer lines and also recombination {\sl continuum} emission.

@@

\subsection{Atomic Data}
Here we use the atomic data described by LK02 and updated by H13. We treat H as a 20-level atom as in SDL05, where 
each level is defined by the principal quantum number, $n$, which is appropriate due to
the degeneracy of each level in the H atom. For the macro-atom treatment of He, 
we have added additional level and line information required 
from \textsc{Topbase} \citep{topbase2005}.  He~\textsc{II} is treated in much the same way as H,
but with 10 levels. He~\textsc{I} has larger energy differentials between different l subshells
and triplet and singlet states. Thus, we still include levels up to $n=10$ but explicitly 
treat the $l$ and $s$ sub-orbitals as distinct levels instead of assuming they are perfectly `l-mixed',
This also allows us to model singlet and 
triplet lines, which are seen in optical CV spectra \citep[e.g.][]{dhillon1996}.

@@

We have previously compared \py to results obtained with  a number of radiative transfer and
photoionization codes. LK02 and H13 conducted comparisons of ionization balance 
with \cld \citep{cloudy2013}, demonstrating excellent agreement. 
We have also carried out comparisons
of ionization and spectral synthesis with the supernova code \textsc{Tardis.} \tar is described by
\cite{kerzendorfsim}, and the spectral comparisons can be found therein.
For the effort reported here, we have carried out tests of
the macro-atom scheme in \py to verify that it does indeed solve for level populations correctly
and produces the expected level emissivities. Figure~\ref{tests} shows good agreement with 
l-mixed Balmer decrement calculations in the Case B limit by \cite{seaton1959}, and 
also compares He \textsc{i} level populations with \textsc{Tardis.}

@@

\py allows for several types of biconical wind models and various options for calculating the spectra of the components of the system.
Here we describe some of the features of the  biconical wind models used in this study of CVs.

@@

For our kinematic wind model we use 
the prescription outlined by Schlosman \& Vitello (1993; hereafter SV93). A schematic is shown
in figure~\ref{cartoon}. In this parametrization, a smooth, biconical
disk wind rises from the accretion disk between radii $r_{min}$ and $r_{max}$. The covering fraction of the wind is 
also controlled by the opening angles of the wind, $\theta_{min}$ and $\theta_{max}$, and the launch angle of
the other streamlines is given by

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/acc_law.eps}
\caption{
The velocity law in our model for various values of
the acceleration exponent, $\alpha$.  {\bf ksl - I am not sure this figure is necessary and do not see where it is discussed.  In light of the fact that our final claims are stated in terms of densisty, it might be interesting to see this expressed as a density rather than a velocity.}
}
\label{acc_law}
\end{figure}

@@

Photons originate from the accretion disk and a central WD in our model. It is also possible to include
a boundary layer with user-defined temperature and emitting area. Emission from the wind is dealt with
by reprocessing of the indivisible energy packets originally produced in the disk or central WD. {\bf ksl -Isn't the wind also allow to emit?  If so, I think saying the wind is dealt with by reporcessing is misleading.}

@@

\py has has some flexibility when treating the accretion disk as a source of photons. 
The disk is broken down into annuli 
such that each annulus contributes an equal amount to the bolometric luminosity. 
We adopt the temperature profile of a standard \cite{shakurasunyaev1973} $\alpha$-disk.
%shown in figure 1.
An annulus can then
be treated either as a blackbody with the corresponding effective temperature or as a stellar atmosphere model
with the appropriate surface gravity and effective temperature. 
Both of these methods have limitations. It is known
from studies of CV accretion disks that they exhibit absorption cores 
formed in an optically thick disk atmosphere.
This can impact the overall spectrum even if a wind is present 
\citep[see e.g.][]{dhillon1996}.
However, the stellar atmosphere models
have had difficulties reproducing the required spectral shape \citep{wade1988}, 
and there are inconsistencies when using a 
stellar atmosphere beneath a dense
wind region that itself goes some way to simulating the disk atmosphere. 
In addition, \cite{linnell2010} found that deviations from the predicted 
$T\propto r^{-3/4}$ temperature profile are required to fit the data
in certain cases.  {\bf ksl - I thought Wade said that both black bodies and stellar atmospheres had trouble reproducing the color of disks.  Please check}

We hope that future simulations, coupled with a comparison to data, will 
actually address some of these problems, particularly since
the filtering effect of the wind may cause noticeable effects
on the colour of the emergent spectrum \citep{hassall}. 
Here, we use a blackbody illuminating spectrum
for the ionization cycles (see figure~\ref{sed}) and to compute our MC estimators.
During the spectral synthesis stage of the simulation we use stellar atmosphere models.
This is partly because they provide the most pessimistic prediction in terms of whether recombination emission in the 
wind can fill in the Balmer jump, due to their intrinsic strong absorption beyond the Balmer edge. {\bf - ksl - This seems like a mistatement.  If we used BB for the detailed spectra there would be no balmer jump problem.}
For low temperatures, we use stellar atmospheres computed 
by Kurucz \citep{kurucz1991}, and for $T>50,000$~K we use models computed with 
\textsc{TLUSTY} \citep{tlusty}. The input spectra are created from these model
atmospheres using \textsc{Synspec}\footnote{http://nova.astro.umd.edu/Synspec43/synspec.html}.
%The temperature profile of our disk,
%with $\dot{M}_{acc} = 10^{-8}~M_{\odot}yr^{-1}$ can be seen in figure~\ref{t_profile}.
% more detail on why we use this hodge podge of atmosphere models?

\subsubsection{Boundary Layer}
The existence of a boundary layer in CVs arises naturally from the boundary condition
of the \cite{shakurasunyaev1973} model, since the material at the inner disk
edge must lose angular momentum in order to be deposited on
the surface of the WD \cite[see e.g.][]{lyndenbell1974}.
Boundary layers are a source of
high energy photons which may be responsible for the photoionization of 
He~\textsc{ii}, so a disk wind can therefore be an important diagnostic
tool for inferring boundary layer luminosities and temperatures. 
The luminosity of the boundary layer, $L_{bl}$, is related to the accretion luminosity
by

@@

It is possible to include a boundary layer in our model which emits 
as a blackbody with a user-specified effective temperature and luminsoity.
In \py, boundary layer photons are then produced around the spherical surface of the 
white dwarf. In our benchmark model, we follow LK02 in setting the boundary layer
luminosity to zero. We investigate the effect
of boundary layers with temperatures between $80{\rm kK}$ and $200{\rm kK}$ 
on the ionization state of the wind and resulting
spectrum in section 5.

@@

As a starting point for our discussion of the effects of a wind on the optical spectra of a CV, we 
take the fiducial model for an SV wind used by LK02 to simulate the UV spectrum of a typical high-state system,
%Our aim here is not to provide a `best fit' model to match any particular dataset. Rather, we takea successful UV wind model as a starting point
and examine the effect of using the better treatment of lines provided
by the macro-atom scheme for H and He.
The model parameters are shown in table 1. 
%they are identical to LK02, and were chosen both to reflect typical
%values for NLs and reproduce realistic UV line profiles.
% The specific model 
% parameters are summarised in table 1, and follow LK02. 
% They are motivated by a need to agree with typical observed values 
% for NLs and the values required to reproduce the UV line profile

@@

Figure~\ref{spec} shows optical synthetic spectra for the benchmark 
model for five inclination angles ranging from $10^\circ$ to $80^\circ$ .
%5 different sightlines.  
The spectra have some lines of H and He which have emission components that are strong enough to overcome
the absorption cores in the underlying disk spectra. These emission lines are generally double peaked, 
as they are formed near the base of the wind where rotational velocity
can still dominate over the slowly increasing poloidal velocity. The line 
opacity is not yet high enough for the poloidal velocity gradient to 
cause single-peaked profiles.  {\bf ksl - So this last statement is based on information that has not yet been submitted as evidence. Be careful of the flow of the paper here.}
We also see a general trend from absorption to emission in spectral lines
with increasing inclination, as one would expect from our wind geometry.
This trend is observed in CVs and can be seen clearly in figure 1. 
% Note, however,
% that in our models the \heiiuv\ line does not exhibit this behaviour, which
% is somewhat surprising. {\bf NOTE: we need to understand this!}.

@@

In addition to examining synthetic spectra, we are also able to 
analyse the physical properties of the outflow. {\bf ksl - A better opening sentence is needed.  We don't really provide much interpretation of this.}
Figure~\ref{wind} shows various physical properties
of the wind geometry. The ionization structure of the wind
is very similar to the original LK02 model, as one might expect.
He is mostly ionized throughout the wind except for a small region near
the disk which is shielded from the photons produced by the hot inner disk.
C\textsc{iv} is the dominant Carbon ion throughout the wind, allowing
for an absorbing column which covers a large range of velocities and is thus
sufficient to produce the required broad, deep and blue-shifted absorption lines
we expect to see in a large fraction of nova-likes.

@@

Clearly, the fiducial model used by LK02 to simulate the UV spectra of CVs does not  explain all
the features seen in the optical spectra of nova-like variables. However,
the wind in this mdoel has a significant effect on features in the synthetic spectrum. 
It suggests that the wind  cannot be ignored as a source of line and continuum emission and as a reprocessing `filter'
of the underlying disk spectral energy distribution (SED).
